---
title: "GEOG 172 Lab 1"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}

#load libraries
library(tidyverse)
library(tibble)
library(dplyr)
library(knitr)
library(here)
library(patchwork)
library(ggrepel)
library(gghighlight)
library(paletteer)
library(ggExtra)
library(ggbeeswarm)
library(ggplot2)
```


# Geog 172: Probability and Statistics (Lab 1)
## Due on October 16, 2024 before 9.30 am

Directions: Please answer all portions of the questions, providing appropriate examples and evidence as needed. Please label the questions you’re trying to answer. Describe the answers
you get in words as well as provide all your calculations. If you use any symbols in your answer
sheet, please state in words what each symbol stands for, and indicate the units of the answers
wherever applicable. Provide the code used to solve these problems with your answers.


*Download the data labeled from Canvas.*

```{r}
incomes <- read_csv(here("data", "median_household_incomes.csv"))
```


**1. The 236 values that appear in the excel sheet on Canvas are the 1990 median household incomes (in dollars) for the 236 census tracts of Buffalo, New York.**

  a) For the first 19 tracts, find the mean, median, range, interquartile range, standard deviation, and variance. In addition, construct a box plot, and a histogram for these 19 observations.
  
```{r}
#make a subset dataframe with just the first 19 rows. 
first19 <- incomes[c(1:19), c(1:2)]

#find the mean
first19_mean <- mean(first19$Income)
#Print results
cat("Mean:", first19_mean, "\n")

#find the median
first19_median <- median(first19$Income)
#Print results
cat("Median:", first19_median, "\n")

#find the range
first19_range <- range(first19$Income)
#Print results
cat("Range:", first19_range, "\n")

#find the interquartile range
first19_iqr <- IQR(first19$Income)
#Print results
cat("Interquartile Range:", first19_iqr, "\n")

#find the standard deviation
first19_sd <- sd(first19$Income)
#Print results
cat("Standard Deviation:", first19_sd, "\n")

#find the variance
first19_var <- var(first19$Income)
#Print results
cat("Variance:", first19_var, "\n")

#construct a boxplot
ggplot(data = first19, aes(x = Income)) +
  geom_boxplot(fill = "seagreen", color = "black") + 
  theme_minimal()

#construct a histogram
ggplot(data = first19, aes(x = Income)) +
  geom_histogram(binwidth = 2000, fill = "coral", color = "black") +
  theme_minimal()

```

  
  b) Repeat part a, this time using all 236 observations.
  
```{r}
#find the mean for ALL observations
all_mean <- mean(incomes$Income)
#Print results
cat("Mean:", all_mean, "\n")

#find the median for ALL observations
all_median <- median(incomes$Income)
#Print results
cat("Median:", all_median, "\n")

#find the range for ALL observations
all_range <- range(incomes$Income)
#Print results
cat("Range:", all_range, "\n")

#find the interquartile range for ALL observations
all_iqr <- IQR(incomes$Income)
#Print results
cat("Interquartile Range:", all_iqr, "\n")

#find the standard deviation for ALL observations
all_sd <- sd(incomes$Income)
#Print results
cat("Standard Deviation:", all_sd, "\n")

#find the variance for ALL observations
all_var <- var(incomes$Income)
#Print results
cat("Variance:", all_var, "\n")

#construct a boxplot for ALL observations
ggplot(data = incomes, aes(x = Income)) +
  geom_boxplot(fill = "seagreen", color = "black") + 
  theme_minimal()

#construct a histogram for ALL observations
ggplot(data = incomes, aes(x = Income)) +
  geom_histogram(binwidth = 5000, fill = "coral", color = "black") +
  theme_minimal()

```

  
  c) Comment on your results. In particular, what does it mean to find the mean of set of medians? How do the observations that have a value of 0 affect the results? Should they be included?
```{r}
#No strict correct answer. Write your thoughts!
```



**2. The following data represent stream link lengths in a river network (given in meters):**

100, 426, 322, 466, 112, 155, 388, 1155, 234, 324, 556, 221, 18, 133, 177, 441.

  Find the mean and standard deviation of the stream lengths.
  
```{r}
stream_link <- c(100, 426, 322, 466, 112, 155, 388, 1155, 234, 324, 556, 221, 18, 133, 177, 441)

#Calculate mean
stream_mean <- mean(stream_link)

#Print results
cat("Mean:", stream_mean, "\n")

#Calculate Standard Deviation
stream_sd <- sd(stream_link)

#Print results
cat("Standard Deviation:", stream_sd, "\n")

```


**3. What is the coefficient of variation among the following commuting times (in minutes):**

23, 43, 42, 7, 23, 11, 23, 55?

```{r}
commuting <- c(23, 43, 42, 7, 23, 11, 23, 55)

# Calculate the mean
mean_value <- mean(commuting)

# Calculate the standard deviation
std_dev <- sd(commuting)

# Calculate the coefficient of variation
cv <- (std_dev / mean_value) * 100

# Print the results
cat("Coefficient of Variation:", cv, "%\n")

```







**4. The following data are derived by Rogerson et al. (1993) from the US National Survey of Families and Households. Cumulative relative frequency distribution of distance to parents among adult children is shown below:**

```{r, echo = FALSE}
families_df <- data.frame(
  Distance = c(5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 
               100, 150, 250, 300, 350, 400, 450, 500, 
               1000, 1500, 2000, 2500, 3000, 3500),
  CumulativeRelativeFrequency = c(0.236, 0.338, 0.418, 0.460, 0.492, 
                                  0.522, 0.546, 0.556, 0.565, 0.575, 
                                  0.643, 0.683, 0.741, 0.756, 0.777, 
                                  0.787, 0.800, 0.808, 0.876, 0.921, 
                                  0.945, 0.966, 0.974, 0.994)
)

# Look at the data frame
kable(families_df, format = "markdown")

```


Questions below apply to adult children who have both parents alive and living together.

  a. What fraction of adult children have parents that live more than 30 miles away,
but no more than 100 miles away?

```{r}
# Calculate the fraction of adult children with parents living more than 30 but no more than 100 miles away
fraction_a <- families_df$CumulativeRelativeFrequency[which(families_df$Distance == 100)] - 
              families_df$CumulativeRelativeFrequency[which(families_df$Distance == 30)]
fraction_a

```

  b. What fraction of adult children has parents that live more than 250 miles away?
  
```{r}
# Calculate the fraction of adult children with parents living more than 250 miles away
fraction_b <- 1 - families_df$CumulativeRelativeFrequency[which(families_df$Distance == 250)]
fraction_b

```
  
  c. What is the likelihood that at least one of three randomly chosen adults will have
parents living more than 100 miles away?

```{r}
# Probability of having parents living more than 100 miles away
p_more_than_100 <- 1 - families_df$CumulativeRelativeFrequency[which(families_df$Distance == 100)]

# Probability that at least one of three has parents living more than 100 miles away
p_at_least_one <- 1 - (1 - p_more_than_100)^3
p_at_least_one

```

  i. Show sample space.
```{r}
sample_space <- c("More than 100 miles", "Not more than 100 miles")
sample_space

```

  ii. Assign probabilities to each element in the sample space.
```{r}
probabilities <- c(p_more_than_100, 1 - p_more_than_100)
names(probabilities) <- sample_space
probabilities

```

  iii. Give the desired probability.
```{r}
#The probability you calculated for at least one of three randomly chosen adults will have parents living more than 100 miles away is already in p_at_least_one.
```


d. What is the median distance that characterizes the spatial separation of parents and their adult children?
```{r}
# Median distance
median_distance <- median(families_df$Distance)
median_distance
```


e. Determine the likelihood that two or more adult children out of a group of ten live more than 1000 miles from their parents.
```{r}
# Probability of living more than 1000 miles away
p_more_than_1000 <- 1 - families_df$CumulativeRelativeFrequency[which(families_df$Distance == 1000)]

# Probability of two or more out of ten
n <- 10
k <- 2
prob_two_or_more <- 1 - pbinom(k - 1, n, p_more_than_1000)
prob_two_or_more

```








**5. Incomes are exponentially distributed with a mean of $10,000. What fraction of the population has income:**
  a) Less than or equal to $8000?
  b) Greater than $12,000?
  c) Between $9,000 and $12,000?
  
```{r}
#Calculating this will involve probability density functions(p.d.f)
# Parameters
mean_income <- 10000

# a. Fraction with income less than or equal to $8000
x_a <- 8000
fraction_a <- 1 - exp(-x_a / mean_income)

# b. Fraction with income greater than $12,000
x_b <- 12000
fraction_b <- exp(-x_b / mean_income)

# c. Fraction with income between $9,000 and $12,000
x_c1 <- 9000
x_c2 <- 12000
fraction_c <- (1 - exp(-x_c2 / mean_income)) - (1 - exp(-x_c1 / mean_income))

# Results
cat("Fraction (a):", fraction_a, "\n")
cat("Fraction (b):", fraction_b, "\n")
cat("Fraction (c):", fraction_c, "\n")

```

  
**6. The number of customers at a bank each day is found to be normally distributed with mean 250 and standard deviation of 110. What fraction of days will have less than 100 customers? More than 320? What number of customers will be exceeded 10% of the time?**

```{r}
# Parameters
mean_customers <- 250
std_dev_customers <- 110

# a. Fraction of days with less than 100 customers
p_less_100 <- pnorm(100, mean = mean_customers, sd = std_dev_customers)

# b. Fraction of days with more than 320 customers
p_more_320 <- 1 - pnorm(320, mean = mean_customers, sd = std_dev_customers)

# c. Number of customers exceeded 10% of the time
# Z-score for 90th percentile
z_90 <- qnorm(0.90)

# Calculate the corresponding number of customers
customers_90th <- mean_customers + z_90 * std_dev_customers

# Results
cat("Fraction of days with less than 100 customers:", p_less_100, "\n")
cat("Fraction of days with more than 320 customers:", p_more_320, "\n")
cat("Number of customers exceeded 10% of the time:", customers_90th, "\n")

```


**7. Assume that the prices paid for housing within a neighborhood have a normal distribution, with mean $100,000, and standard deviation $35,000**

  a. What percentage of houses in the neighborhood have prices between
$90,000 and $130,000?
  b. What price of housing is such that only 12% of all houses in the
neighborhood have lower prices?

```{r}
# Parameters
mean_price <- 100000
std_dev_price <- 35000

# a. Percentage of houses with prices between $90,000 and $130,000
p_between_90k_130k <- pnorm(130000, mean = mean_price, sd = std_dev_price) - 
                       pnorm(90000, mean = mean_price, sd = std_dev_price)

# Convert to percentage
percentage_between_90k_130k <- p_between_90k_130k * 100

# b. Price such that only 12% of houses have lower prices
z_12 <- qnorm(0.12)
price_12_percent <- mean_price + z_12 * std_dev_price

# Results
cat("Percentage of houses between $90,000 and $130,000:", percentage_between_90k_130k, "%\n")
cat("Price of housing such that only 12% have lower prices:", price_12_percent, "\n")

```


**8. If the probability that an individual moves outside of their country of residence in a given year is 0.15, what is the probability that**
  a. Less than three out of a sample of ten move outside the country?
  b. At least one moves outside the country?
  
```{r}
# Parameters
p <- 0.15  # probability of moving
n <- 10    # sample size

# a. Probability that less than 3 move outside the country
p_less_than_3 <- sum(dbinom(0:2, size = n, prob = p))

# b. Probability that at least one moves outside the country
p_at_least_one <- 1 - dbinom(0, size = n, prob = p)

# Results
cat("Probability that less than 3 move outside the country:", p_less_than_3, "\n")
cat("Probability that at least one moves outside the country:", p_at_least_one, "\n")

```


**9. There are an average of four accidents per month at an intersection.**
  a. What is the probability of three or more accidents next month?
  b. What is the probability of two or more accidents in a 23-day period

```{r}
# Parameters
lambda_month <- 4              # average accidents per month
lambda_23_days <- (4 / 30) * 23 # average accidents in 23 days

# a. Probability of three or more accidents next month
p_three_or_more <- 1 - (dpois(0, lambda_month) + 
                         dpois(1, lambda_month) + 
                         dpois(2, lambda_month))

# b. Probability of two or more accidents in a 23-day period
p_two_or_more_23_days <- 1 - (dpois(0, lambda_23_days) + 
                               dpois(1, lambda_23_days))

# Results
cat("Probability of three or more accidents next month:", p_three_or_more, "\n")
cat("Probability of two or more accidents in a 23-day period:", p_two_or_more_23_days, "\n")

```


**10. The annual probability of moving is 0.17.**

  a. What is the probability that a person moves for the first time in the first year of observation?
  
  b. What is the probability that the person moves for the first time in the fourth year of observation?
  
  c. What is the mean time it takes for an individual to move?
  
  d. What is the variance of the time it takes for a person to move?
  
  e. Among ten people, what is the probability that three of fewer move in a given year?
  
```{r}
# Parameters
p <- 0.17  # probability of moving
q <- 1 - p # probability of not moving
n <- 10    # sample size

# a. Probability of moving for the first time in the first year
prob_first_year <- p

# b. Probability of moving for the first time in the fourth year
prob_fourth_year <- q^3 * p

# c. Mean time it takes for an individual to move
mean_time <- 1 / p

# d. Variance of the time it takes for a person to move
variance_time <- q / (p^2)

# e. Probability that three or fewer move among ten people
p_three_or_fewer <- sum(dbinom(0:3, size = n, prob = p))

# Results
cat("Probability of moving for the first time in the first year:", prob_first_year, "\n")
cat("Probability of moving for the first time in the fourth year:", prob_fourth_year, "\n")
cat("Mean time to move:", mean_time, "\n")
cat("Variance of time to move:", variance_time, "\n")
cat("Probability that three or fewer move among ten people:", p_three_or_fewer, "\n")

```

